Short and Mechanized Logical Relation for Dependent Type Theory

### 2024-02-23 paper/paper.tex
comments on what we still need to say in the first three sections

Our goal with this work is to demonstrate how the proof technique of
\emph{logical relations} can be applied to reason about dependent type theories.
Tutorial material on this topic~\cite{lr-tutorials} is primarily focussed on
systems with simple or polymorphic types. In that context, logical relations
can be define as simple recursive functions over the structure of types, or
(in the case of recursive types) defined over the evaluation steps of the
computation.

@@

@@

\section{Logical Relation}
% \scw{This section pre-supposes that we want to define a logical relation,
% but doesn't precisely state what we want to use it to prove, and why a logical
% relation is suitable. (And why the property you want to prove is difficult to
% show!) Should add more motivation here.}
Before we define our logical relation, we first formally specify the
consistency property that we want to prove.
\begin{theorem}[Logical Consistency]
  \label{theorem:consistency}
  The judgment $[[empty |- a : Void ]]$ is not derivable.
\end{theorem}
The property can be formulated in a simply typed language, where
$[[Void]]$ is similarly defined as a type that has no term. A related
property, referred to as the termination property (for closed terms),
is commonly used in introductory materials such as
\citet{skorstengaard2019introduction}, \citet{pierce2002types}, and
\citet{harpertait} to motivate the need for a logical relation.

>>>>

Our goal with this work is to demonstrate how the proof technique of
\emph{logical relations} can be applied to reason about dependent type theories.
Tutorial material on this topic~\cite{lr-tutorials} is primarily focussed on
systems with simple or polymorphic types. In that context, logical relations
can be define as simple recursive functions over the structure of types, or
(in the case of recursive types) defined over the evaluation steps of the
computation.
\scw{need one sentence to motivate why logical relations are important}

@@

\scw{This section needs to say:
  \begin{itemize}
\item definitional equality is untyped so that we can prove 
  properties about it independent of the type system
\item definitional equality is based on parallel reduction so that we can take
  advantage of the algorithmic structure later. This definition is easier to
  invert because there are fewer derivation. That is important when reasoning
  about our logical relation.
\item can prove that parallel reduction is equivalent to other definitions of
  equality using standard means (cite Barendregt's book).  (we go halfway
  there by showing it is an equivalence relation). Can also prove equivalent
  to typed relation. (cite Siles, also CoreSpec?)
\item All of the proofs in this section are standard, well known, and use
  techniques that are well-suited for proof assistants. In fact, de Bruijn's
  paper that introduces de Bruijn indices was part of a mechanized confluence
  proof for the untyped lambda calculus.
\item extensions: forward reference to eta equivalence for functions in later
  section. type-directed equivalence is future work
  \end{itemize}
}

@@

\section{Logical Relation}
% \scw{This section pre-supposes that we want to define a logical relation,
% but doesn't precisely state what we want to use it to prove, and why a logical
% relation is suitable. (And why the property you want to prove is difficult to
% show!) Should add more motivation here.}
\scw{We need to explicitly point out that the key ideas of this paper are 
  discussed, here, in this section. 
  We need to explicitly remark on why logical relations are difficult to 
  define for dependent type theory and explain why this setting is more 
  difficult than with simple types (STLC) or with polymorphic types (System F).
  \begin{itemize}
  \item Dependent function types (result type mentions argument value, so can't define 
    relation by recursion over type structure)
  \item Definitional equality (not all types look like types)
  \item Identity types(?)
  \end{itemize}
  Should we be more explicit in our comparison with Girard's trick for polymorphic type?
  There, the definition stays recursive because it doesn't substitute for the variables 
  in the function types. But that approach is not available in this setting, because not 
  all quantified things are types. And we might need that information to interpret, say, 
  identity types in the right way.
}
\scw{ We also need to explicitly point out that our logical relation is untyped. 
  This has two benefits: it allows semantic typing to be meaningful independent from 
  syntactic typing (cite Derek, forward reference to next section) and it avoids 
  significant bookkeeping, especially in the case of Kripke logical relations (we need to define 
  what these are).
  Is there a cost to an untyped relation? 
}
Before we define our logical relation, we first formally specify the
consistency property that we want to prove.
\begin{theorem}[Logical Consistency]
  \label{theorem:consistency}
  The judgment $[[empty |- a : Void ]]$ is not derivable.
\end{theorem}
The property can be formulated in a simply typed language, where
$[[Void]]$ is similarly defined as a type that has no term. A related
property, referred to as the termination property (for closed terms),
is commonly used in introductory materials such as
\citet{skorstengaard2019introduction}, \citet{pierce2002types}, and
\citet{harpertait} to motivate the need for a logical relation.
